### 2018-03-07 submission/manuscript.tex
incorporate comments from Pat on results section

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased
Odds Ratio (OR) of Tumors:}} We first assessed whether variation in
broad community metrics like total number of operational taxonomic units
(OTUs) (i.e.~richness), the evenness of their abundance, and the overall
diversity was associated with disease stage after controlling for study
and variable region differences. In stool, there was a significant
decrease in both evenness and diversity as disease severity progressed
from normal to adenoma to carcinoma (P-value = 0.025 and 0.043,
respectively) {[}Figure 1{]}; there was not a significant difference for
richness (P-value = 0.21). We next tested whether the decrease in these
community metrics translated into significant ORs for having an adenoma
or carcinoma. For fecal samples, the ORs for richness were not
significantly greater than 1.0 for adenoma or carcinoma (P-value = 0.40)
{[}Figure 2A{]}. The ORs for evenness were significantly higher than 1.0
for adenoma (OR = 1.3 (1.02 - 1.65), P-value = 0.035) and carcinoma (OR
= 1.66 (1.2 - 2.3), P-value = 0.0021) {[}Figure 2B{]}. The ORs for
diversity were only significantly greater than 1.0 for carcinoma (OR =
1.61 (1.14 - 2.28), P-value = 0.0069), but not for adenoma (P-value =
0.11) {[}Figure 2C{]}. Although these OR are significantly greater than
1.0, it is doubtful that these are clinically meaningful values.

@@

\textbf{\emph{Disease Progression is Associated with Community-Wide
Changes in Composition and Abundance:}} Based on the differences in
evenness and diversity, we next asked whether there were community-wide
differences in the structure of the communities associated with
different disease stages. We identified significant bacterial community
differences in the stool of patients with adenomas relative to those
with normal colons in 1 of 4 studies and in patients with carcinomas
relative to those with normal colons in 6 of 7 studies (PERMANOVA;
P-value \textless{} 0.05) {[}Table S2{]}. Similar to the analyses using
stool samples, there were significant differences in bacterial community
structure between subjects with normal colons and those with adenoma (1
of 2 studies) and carcinoma (1 of 3 studies) {[}Table S2{]}. For studies
that used matched samples no differences in bacterial community
structures were observed {[}Table S2{]}. Combined, these results
indicate that there consistent and significant community-wide changes in
the fecal community structure of subjects with carcinomas. However, the
signal observed in subjects with adenomas or when using tissue samples
was not as consistent. This is likely due to a smaller effect size or
the relatively small sample sizes among the studies that characterized
the tissue microbiota.

\textbf{\emph{Individual Taxa are Associated with Significant ORs for
Carcinomas:}} Next we identified those taxa were associated with ORs
that were significantly associated with having a normal colon or the
presence of adenomas or carcinomas. No taxa had a significant OR for the
presence of adenomas when we used data collected from stool or tissue
samples (Table S3 \& S4). In contrast, 8 taxa had significant ORs for
the presence of carcinomas using data from stool samples. Of these, 4
are commonly associated with the oral cavity: Fusobacterium (OR = 2.74
(1.95 - 3.85)), Parvimonas (OR = 3.07 (2.11 - 4.46)), Porphyromonas (OR
= 3.2 (2.26 - 4.54)), and Peptostreptococcus (OR = 7.11 (3.84 - 13.17))
{[}Table S3{]}. The other 4 were Clostridium XI (OR = 0.65 (0.49 -
0.86)), Enterobacteriaceae (OR = 1.79 (1.33 - 2.41)), Escherichia (OR =
2.15 (1.57 - 2.95)), and Ruminococcus (OR = 0.63 (0.48 - 0.83)). Among
the data collected from tissue samples, only unmatched carcinoma samples
had taxa with a significant OR. Those included Dorea (OR = 0.35 (0.22 -
0.55)), Blautia (OR = 0.47 (0.3 - 0.73)), and Weissella (OR = 5.15 (2.02
- 13.14)). Mouth-associated genera were not significantly associated
with an increased OR for carcinoma in tissue samples {[}Table S4{]}. For
example, Fusobacterium had an OR of 3.98 (1.19 - 13.24; however, due to
the small number of studies and considerable variation in the data, the
Benjimani-Hochberg-corrected P-value was 0.93 {[}Table S4{]}. It is
interesting to note that Ruminococcus and members of Clostridium group
XI in stool and Dorea and Blautia in tissue had ORs that were
significantly less than 1.0, which suggests that these populations are
protective against the development of carcinomas. Overall, there was no
overlap in the taxa with significant OR between stool and tissue
samples.

\textbf{\emph{Individual taxa with a significant OR do a poor job of
differentiating subjects with normal colons and those with carcinoma:}}
We next asked whether those taxa that had a significant OR associated
with having a normal colon or carcinomas could be used individually, to
classify subjects as having a normal colon or carcinomas. Whereas the OR
was defined based on whether the relative abundance for a taxon in a
subject was above or below the median relative abundance for that taxon
across all subjects in a study, we generated receiver operator
characteristic (ROC) curves for each taxon in each study and calculated
the area under the curve (AUC). This allowed us to use a more fluid
relative abundance threshold for defining disease status. Using data
from stool samples, the 8 taxa did no better at classifying the subjects
than one would expect by chance (i.e.~AUC=0.50) {[}Figure 3A{]}. The
taxa that performed the best included Clostridium XI, Ruminococcus, and
Escherichia and even these had median AUC values less than 0.588.
Likewise, in unmatched tissue samples the 8 taxa with significant ORs
taxa were marginally better than one would expect by chance {[}Figure
3B{]}. The relative abundance of Dorea was the best predictor of
carcinomas and its median AUC was only 0.62. These results suggest that
although these taxa are associated with a decreased or increased OR for
the presences of carcinomas, individually, they do a poor job of
classifying a subject's disease status.

\textbf{\emph{Combined taxa model classifies subjects better than using
individual taxa:}} Instead of attempting to classify subjects based on
individual taxa, next we generated Random Forest models that combined
the individual taxa and evaluated the ability to classify as subject's
disease status. For data from stool samples, the combined model had an
AUC of 0.75, which was significantly higher than any of the AUC values
for the individual taxa (P-value \textless{} 0.033). For the full taxa
models using stool, \emph{Bacteroides} and \emph{Lachnospiraceae} were
the most common taxa in the top 10\% mean decrease in accuracy (MDA)
across studies {[}Figure S2{]}. Similarly, using data from the unmatched
tissue samples, the combined model had an AUC of 0.77, which was
significantly higher than the AUC values for Blautia and Weissella
(P-value \textless{} 0.037). For the full taxa models using unmatched
tissue, \emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
\emph{Ruminococcaceae} were the most common taxa in the top 10\% mean
decrease in accuracy across studies {[}Figure S3{]}. Clearly, pooling
the information from the taxa with significant ORs results in a model
that outperforms classifications made using individual taxa.

\textbf{\emph{Performance of models based on taxa relative abundance in
full community are better than those based on taxa with significant
ORs:}} Next, we asked whether a Random Forest classification model built
using all of the taxa found in the communities would outperform the
models generated using those taxa with a significant OR. Similar to our
inability to identify taxa associated with a significant OR for the
presence of adenomas, the median AUCs to classify subjects as having
normal colons or having adenomas using data from stool or tissue samples
were marginally better than 0.5 for any study{[}Figure 4A \& S4A{]}. In
contrast, the models for classifying subjects as having normal colons or
having carcinomas using data from stool or tissue samples yielded AUC
values meaningfully higher than 0.5 {[}Figure 4B \& S4B-C{]}. When we
compared the models based on all of the taxa in a community to models
based on the taxa with significant ORs, the results were mixed. Using
the data from stool samples we found that although the AUC for 6 of 7
studies increased (mean decrease = 9.53\%), the more expansive models
performed worse for 1 of the studies (decrease = 0.38\%). The overall
improvement in performance was statistically significant (one-tailed
paired T-test; P-value = 0.005). Of the 8 taxa with significant ORs, all
8 were among the top 10\% most important taxa as measured by mean
decrease in accuracy, in at least one study . Similarly, using the data
from unmatched tissue samples we found that the AUC for 4 out of 4
studies decreased between full versus select OR models (mean decrease =
19.11\%, one-tailed paired T-test; P-value = 0.03). Of the 3 taxa with
significant ORs, all 3 were among the top 10\% most important taxa as
measured by mean decrease in accuracy, in at least one study. The most
important taxa across study within the significant OR taxa only models
for stool were \emph{Ruminococcus} and \emph{Clostridium XI} {[}Figure
5A{]}. For the singificant OR taxa unmatched tissue models both
\emph{Dorea} and \emph{Blautia} were the important based on mean
decrease in accuracy {[}Figure 5B{]}. These results were surprising
because it demonstrated that the ability to classify subjects could be
done based on a limited characterization of the communities.

\textbf{\emph{Performance of models based on OTU relative abundance in
full community are not significantly better than those based on taxa
with significant ORs:}} The previous models were based on relative
abundance data where sequences were assigned to coarse taxonomic
assignments (i.e.~typically genus or family level). To determine whether
model performance improved with a more fine scale classification, we
assigned sequences to operational taxonomic units (OTUs) where the
similarity among sequences within an OTU was more than 97\%. We again
found that classification models built using all of the sequence data
for a community did a poor job of differentiating between subjects with
normal colons and those with adenomas (median AUC: 0.53 {[}0.37-
0.56{]}), but did a good job of differentiating between subjects with
normal colons and those with carcinomas (median AUC: 0.71 {[}0.5-
0.9{]}). The OTU-based models performed similarly to those constructed
using the taxa with significant ORs (one-tailed paired T-test; P-value =
0.966) and those using all taxa (one-tailed paired T-test; P-value =
0.146). Among the OTUs that had the highest mean decrease in accuracy
for the OTU-based models, we found that OTUs that affiliated with all of
the 8 taxa that had a significant OR were within the top 10\% for at
least one study. Again, this result was surprising as it indicated that
a finer scale classification of the sequence data and thus a larger
number of features to select from, did not yield improved classification
of the subjects.

\textbf{\emph{Generalizability of taxon-based models trained on one
dataset to the other datasets:}} Considering the good performance of the
Random Forest models using taxa with a significant OR and using all of
the taxa, we next asked how well the models would perform when given
data from a different subject cohort. For instance, if a model was
trained using data from the Ahn study, we wanted to know how well it
would perform using the data from the Baxter study. We found the models
trained using the taxa with a significant OR all had a higher median AUC
than the models trained using all of the taxa when tested on the other
datasets {[}Figure 6 \& S5{]}. As might be expected, the difference
between the performance of the modelling approaches appeared to vary
with the size of the training cohort {[}Figure 6{]}. These data suggest
that given a sufficient number of subjects with normal colons and
carcinomas, Random Forest models trained using a small number of taxa
can accurately classify individuals from a different cohort.

>>>>

\textbf{\emph{Lower bacterial diversity is associated with increased
odds ratio (OR) of tumors:}} We first assessed whether variation in
broad community metrics like total number of operational taxonomic units
(OTUs) (i.e.~richness), the evenness of their abundance, and the overall
diversity was associated with disease stage after controlling for study
and variable region differences. In stool, there was a significant
decrease in both evenness and diversity as disease severity progressed
from normal to adenoma to carcinoma (P-value = 0.025 and 0.043,
respectively) {[}Figure 1{]}; there was not a significant difference for
richness (P-value = 0.21). We next tested whether the decrease in these
community metrics translated into significant ORs for having an adenoma
or carcinoma. For fecal samples, the ORs for richness were not
significantly greater than 1.0 for adenoma or carcinoma (P-value = 0.40)
{[}Figure 2A{]}. The ORs for evenness were significantly higher than 1.0
for adenoma (OR = 1.3 (1.02 - 1.65), P-value = 0.035) and carcinoma (OR
= 1.66 (1.2 - 2.3), P-value = 0.0021) {[}Figure 2B{]}. The ORs for
diversity were only significantly greater than 1.0 for carcinoma (OR =
1.61 (1.14 - 2.28), P-value = 0.0069), but not for adenoma (P-value =
0.11) {[}Figure 2C{]}. Although these OR are significantly greater than
1.0, it is doubtful that these are clinically meaningful values.

@@

\textbf{\emph{Disease progression is associated with community-wide
changes in composition and abundance:}} Based on the differences in
evenness and diversity, we next asked whether there were community-wide
differences in the structure of the communities associated with
different disease stages. We identified significant bacterial community
differences in the stool of patients with adenomas relative to those
with normal colons in 1 of 4 studies and in patients with carcinomas
relative to those with normal colons in 6 of 7 studies (PERMANOVA;
P-value \textless{} 0.05) {[}Table S2{]}. Similar to the analyses using
stool samples, there were significant differences in bacterial community
structure between subjects with normal colons and those with adenoma (1
of 2 studies) and carcinoma (1 of 3 studies) {[}Table S2{]}. For studies
that used matched samples no differences in bacterial community
structures were observed {[}Table S2{]}. Combined, these results
indicate that there consistent and significant community-wide changes in
the fecal community structure of subjects with carcinomas. However, the
signal observed in subjects with adenomas or when using tissue samples
was not as consistent. This is likely due to a smaller effect size or
the relatively small sample sizes among the studies that characterized
the tissue microbiota.

\textbf{\emph{Individual taxa are associated with significant ORs for
carcinomas:}} Next we identified those taxa were associated with ORs
that were significantly associated with having a normal colon or the
presence of adenomas or carcinomas. No taxa had a significant OR for the
presence of adenomas when we used data collected from stool or tissue
samples (Table S3 \& S4). In contrast, 8 taxa had significant ORs for
the presence of carcinomas using data from stool samples. Of these, 4
are commonly associated with the oral cavity: \emph{Fusobacterium} (OR =
2.74 (1.95 - 3.85)), \emph{Parvimonas} (OR = 3.07 (2.11 - 4.46)),
\emph{Porphyromonas} (OR = 3.2 (2.26 - 4.54)), and
\emph{Peptostreptococcus} (OR = 7.11 (3.84 - 13.17)) {[}Table S3{]}. The
other 4 were \emph{Clostridium XI} (OR = 0.65 (0.49 - 0.86)),
\emph{Enterobacteriaceae} (OR = 1.79 (1.33 - 2.41)), \emph{Escherichia}
(OR = 2.15 (1.57 - 2.95)), and \emph{Ruminococcus} (OR = 0.63 (0.48 -
0.83)). Among the data collected from tissue samples, only unmatched
carcinoma samples had taxa with a significant OR. Those included
\emph{Dorea} (OR = 0.35 (0.22 - 0.55)), \emph{Blautia} (OR = 0.47 (0.3 -
0.73)), and \emph{Weissella} (OR = 5.15 (2.02 - 13.14)).
Mouth-associated genera were not significantly associated with an
increased OR for carcinoma in tissue samples {[}Table S4{]}. For
example, \emph{Fusobacterium} had an OR of 3.98 (1.19 - 13.24; however,
due to the small number of studies and considerable variation in the
data, the Benjimani-Hochberg-corrected P-value was 0.93 {[}Table S4{]}.
It is interesting to note that \emph{Ruminococcus} and members of
\emph{Clostridium group XI} in stool and \emph{Dorea} and \emph{Blautia}
in tissue had ORs that were significantly less than 1.0, which suggests
that these populations are protective against the development of
carcinomas. Overall, there was no overlap in the taxa with significant
OR between stool and tissue samples.

\textbf{\emph{Individual taxa with a significant OR do a poor job of
differentiating subjects with normal colons and those with carcinoma:}}
We next asked whether those taxa that had a significant OR associated
with having a normal colon or carcinomas could be used individually, to
classify subjects as having a normal colon or carcinomas. OR values were
caluclated based on whether the relative abundance for a taxon in a
subject was above or below the median relative abundance for that taxon
across all subjects in a study. To measure the ability of these taxa to
classify individuals we instead generated receiver operator
characteristic (ROC) curves for each taxon in each study and calculated
the area under the curve (AUC). This allowed us to use a more fluid
relative abundance threshold for defining disease status. Using data
from stool samples, the 8 taxa did no better at classifying the subjects
than one would expect by chance (i.e.~AUC=0.50) {[}Figure 3A{]}. The
taxa that performed the best included \emph{Clostridium XI},
\emph{Ruminococcus}, and \emph{Escherichia}, however, these had median
AUC values less than 0.588. Likewise, in unmatched tissue samples the 8
taxa with significant ORs taxa were marginally better than one would
expect by chance {[}Figure 3B{]}. The relative abundance of Dorea was
the best predictor of carcinomas and its median AUC was only 0.62. These
results suggest that although these taxa are associated with a decreased
or increased OR for the presences of carcinomas, individually, they do a
poor job of classifying a subject's disease status.

\textbf{\emph{Combined taxa model classifies subjects better than using
individual taxa:}} Instead of attempting to classify subjects based on
individual taxa, next we combined information from the individual taxa
and evaluated the ability to classify a subject's disease status using
Random Forest models. For data from stool samples, the combined model
had an AUC of 0.75, which was significantly higher than any of the AUC
values for the individual taxa (P-value \textless{} 0.033). For the full
taxa models using stool, \emph{Bacteroides} and \emph{Lachnospiraceae}
were the most common taxa in the top 10\% mean decrease in accuracy
(MDA) across studies {[}Figure S2{]}. Of the 3 taxa with significant
ORs, all 3 were among the top 10\% most important taxa as measured by
mean decrease in accuracy, in at least one study. The most important
taxa across study within the significant OR taxa only models for stool
were \emph{Ruminococcus} and \emph{Clostridium XI} {[}Figure 5A{]}.
Similarly, using data from the unmatched tissue samples, the combined
model had an AUC of 0.77, which was significantly higher than the AUC
values for Blautia and Weissella (P-value \textless{} 0.037). For the
full taxa models using unmatched tissue, \emph{Lachnospiraceae},
\emph{Bacteroidaceae}, and \emph{Ruminococcaceae} were the most common
taxa in the top 10\% mean decrease in accuracy across studies {[}Figure
S3{]}. For the singificant OR taxa unmatched tissue models both
\emph{Dorea} and \emph{Blautia} were the important based on mean
decrease in accuracy {[}Figure 5B{]}. Pooling the information from the
taxa with significant ORs results in a model that outperforms
classifications made using individual taxa.

\textbf{\emph{Performance of models based on taxa relative abundance in
full community are better than those based on taxa with significant
ORs:}} Next, we asked whether a Random Forest classification model built
using all of the taxa found in the communities would outperform the
models generated using those taxa with a significant OR. Similar to our
inability to identify taxa associated with a significant OR for the
presence of adenomas, the median AUCs to classify subjects as having
normal colons or having adenomas using data from stool or tissue samples
were only marginally better than 0.5 for any study (median AUC =
0.5486298 {[}0.3671667 - 0.971{]}) {[}Figure 4A \& S4A{]}. In contrast,
the models for classifying subjects as having normal colons or having
carcinomas using data from stool or tissue samples yielded AUC values
meaningfully higher than 0.5 {[}Figure 4B \& S4B-C{]}. When we compared
the models based on all of the taxa in a community to models based on
the taxa with significant ORs, the results were mixed. Using the data
from stool samples we found that the AUC for 6 of 7 studies increased by
an average of -14.8\%) and AUC for the Flemer study decreased by
0.54\%). The overall improvement in performance was statistically
significant (mean = 12.61\%, one-tailed paired T-test; P-value = 0.005).
Similarly, using the data from unmatched tissue samples we found that
the AUC of studies increased by an average of 19.11\% when we used all
of the taxa rather than the 3 taxa with significant ORs (one-tailed
paired T-test; P-value = 0.03). Although the significant OR taxa models
can classify indidivuals with and without carcinoma tumors, they are
still missing taxa from the full community models that can increase the
model accuracy.

\textbf{\emph{Performance of models based on OTU relative abundance in
full community are not significantly better than those based on taxa
with significant ORs:}} The previous models were based on relative
abundance data where sequences were classified to coarse taxonomic
assignments (i.e.~typically genus or family level). To determine whether
model performance improved with finer scale classification, we assigned
sequences to operational taxonomic units (OTUs) where the similarity
among sequences within an OTU was more than 97\%. We again found that
classification models built using all of the sequence data for a
community did a poor job of differentiating between subjects with normal
colons and those with adenomas (median AUC: 0.53 {[}0.37- 0.56{]}), but
did a good job of differentiating between subjects with normal colons
and those with carcinomas (median AUC: 0.71 {[}0.5- 0.9{]}). The
OTU-based models performed similarly to those constructed using the taxa
with significant ORs (one-tailed paired T-test; P-value = 0.966) and
those using all taxa (one-tailed paired T-test; P-value = 0.146)
{[}Figure 4{]}. Among the OTUs that had the highest mean decrease in
accuracy for the OTU-based models, we found that OTUs that affiliated
with all of the 8 taxa that had a significant OR were within the top
10\% for at least one study. This result was surprising as it indicated
that a finer scale classification of the sequences and thus a larger
number of features to select from, did not yield improved classification
of the subjects.

\textbf{\emph{Generalizability of taxon-based models trained on one
dataset to the other datasets:}} Considering the good performance of the
Random Forest models using taxa with a significant OR and using all of
the taxa, we next asked how well the models would perform when given
data from a different subject cohort. For instance, if a model was
trained using data from the Ahn study, we wanted to know how well it
would perform using the data from the Baxter study. We found the models
trained using the taxa with a significant OR all had a higher median AUC
than the models trained using all of the taxa when tested on the other
datasets {[}Figure 6 \& S5{]}. As might be expected, the difference
between the performance of the modelling approaches appeared to vary
with the size of the training cohort (R\textsuperscript{2} = 0.66)
{[}Figure 6{]}. These data suggest that given a sufficient number of
subjects with normal colons and carcinomas, Random Forest models trained
using a small number of taxa can accurately classify individuals from a
different cohort.
